---
title: "MICB425 portfolio"
author: "Jessica Ngo (31837131)"
date: "version `r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_float:
      collapsed: no
    theme: united
---
# Evidence worksheets

```{r child = "evidencePS/Evidence_worksheet_01.Rmd"}

```

```{r child = "evidencePS/Evidence_worksheet_02.Rmd"}

```

```{r child = "evidencePS/Evidence_worksheet_03.Rmd"}

```

```{r child = "evidencePS/Evidence_worksheet_04.Rmd"}

```

```{r child = "evidencePS/Evidence_worksheet_05.Rmd"}

```

# Problem sets

```{r child = "evidencePS/Problem_set_01.Rmd"}

```

```{r child = "evidencePS/Problem_set_02.Rmd"}

```

```{r child = "evidencePS/Problem_set_03.Rmd"}

```

```{r child = "evidencePS/Problem_set_04.Rmd"}

```

# Writing assignments

## Writing assignment Module 1
*Prompt: "Microbial life can easily live without us; we, however, cannot survive without the global catalysis and environmental transformations it provides." Do you agree or disagree with this statement? Answer the question using specific reference to your reading, discussions and content from evidence worksheets and problem sets. * 

**Introduction**  

"Microbial life can easily live without us; we, however, cannot survive without the global catalysis and environmental transformations it provides." Historically, microbes have existed and fared well long before the existence of humans, with fossils dating back to 0.2 million years ago (Ma) [1], and probably existing before 3500 Ma [2]. However, for humans, I believe that we are unable to live without the global catalysis and environmental transformations that they provide, while microbial life is probably content to go on without us in whatever state we put the world in. In this paper, the microbes' irreplaceable and essential role in supporting human life will be examined. This involves discussion of why microbes would be impossible to replace with human technology, both in quantity (abundance, catalysis on a global scale) and quality (complexity).   

**Quantity**  

Firstly, simply due to the sheer number and abundance of microbes, microbes would be extremely difficult or impossible to replace. While it is common knowledge that microbial life is everywhere, quantitative classification is still largely disregarded. Based on the work of Whitman, Coleman, and Wiebe in 1998, the number provided for prokaryotic cells on Earth was approximated to be 4-6 x 1030 cells, which is responsible for 350-550Pg of carbon [3]. This number is so massive, it is easy to trivialize, since we generally do not have a concept of this number. To put this in perspective, consider that if we were to count every grain of sand on earth in all deserts and beaches, which would be roughly equal 7.5 x 1018 grains of sand. This does not even come close to the number of prokaryotic cells on earth. However this might have been obvious, considering the fact that microbes are everywhere and also magnitudes smaller in size. More comparably, the number of stars in the universe has been approximated to be 7 x 1022. This still does not come close to the number of prokaryotic cells. Each one of these cells is constantly under natural selection pressures, evolving, and adapting to our ever changing climate, responsible for catalyzing reactions, providing energy as a primary producer, and acting as a huge storage of all of the information used to carry out all these roles and processes. Therefore, to begin considering a life without microbes, or replacing microbes, we would have to somehow be able to do the equivalent work of 4-6 x 1030 cells, a number that we cannot even realistically conceptualize.  
    
Even if we were to put the microbes' role as a large primary producer aside, that still leaves the role microbes play in the cycling of elements on a global scale. In particular, Nitrogen, which is essential for synthesis of nucleic acids and proteins, and Carbon, an essential element for all organic life.  In the name of monetary profit, and convenience, humans have made a huge impact on the Nitrogen cycle in the 20th century. This includes the reduction of the inert form of Nitrogen, N2 to NH4+, the implementation of new agricultural practices to boost yields, and burning of fossil fuels [3].  

In contrast to these technological advances, very little attempt has been made towards keeping the natural order and balance of Nitrogen, or other elements in general. As a result of this misuse, there have been significant effects to the environment, such as nitrogen loss and subsequent eutrophication of coastal waters - where diversity of life is dying out due to the newly created hypoxic environments [3].  

As indicated by this detrimental effect of humans on the Nitrogen cycle, humans are in fact capable of manipulating the cycling of elements on a global scale, just like the microbes. However, unlike the microbes, humans in general do not care for what state an element is in unless it is profitable to do so. This means that only the elements that we deem "useful" will be kept in supply, while others will be in whatever form or state is most convenient. This also means that the balance of elements, or preserving of the cycles, including the Nitrogen cycle, or any other cycle is of no concern at all. There is no respect, or "land ethic" that frowns upon the most efficient commercial processes that put elements out of balance and in disarray instead. Only the microbes are concerned about keeping a balance, since they are not motivated by profits, but rather, by survival. As they grow and adapt to niches, an equilibrium is achieved. If not for the microbes' essential role in cycling these elements, we would already be out of Nitrogen and extinct.  

An even more alarming example of element consumption is Carbon. The carbon cycle has a "leak" where a small amount of organic carbon sinks to the ocean floor and becomes part of sediment, and is unable to re-enter the cycle [4]. This leak, paired with the industrialization of the human race and burning of fossil fuels has resulted in massive emissions of carbon dioxide into the atmosphere, which could potentially result in a mass extinction within the next century [4]. In consideration of the coordination required to alter the both the nitrogen and carbon cycles (ie. All of humanity's fertilization of crops, and burning fossil fuels), it is highly unlikely that the human race would be able to concert a coordinated effort towards innovation and preserving the balance of the different states of the elements until it is too late. If we are unable to sacrifice financial profit to replace even an aspect of the microbes role in element cycling to prevent extinction, it is impossible to even begin considering replacing the microbes role in cycling every single element on Earth.  

**Quality**  

Even if humans were somehow miraculously united and agreed to put all resources towards sustainable practices and research to maintain the balance of the mentioned elements, the complex roles that microbes play in an interconnected ecological web still remain to be understood. After all, how can we replace something if we don't even know what it does, or how it works? Through trial and error, evolution has given us an order of things that is sustainable, yet we are throwing things out of balance, and not stopping. Additionally, in a naïve attempt to fix some of our wrongdoings, we have made things worse, due to a lack of understanding of the complexity involved with a problem. For example, to manage the invasive cane toads originally brought to Australia, the Lungworms were introduced, since the introduced cane toads were overrunning the native tree frog population [5]. Unfortunately, lungworms also ended up killing the tree frog numbers as well [5]. It is wholly possible that humans trying to fix the cycling of elements on a global scale by counterbalancing with another human-introduced intervention could actually make things worse than before, similar to the cane toad example. Additionally, microbes are also involved in their own unique communities and part of every single complex ecological web on earth. Trying to replace the microbes in these roles without a complete understanding would be a mistake, and a complete gamble. Everything is interconnected, and human research has not yet come to a mature enough level to even begin understanding the intricacies and nuances of the delicate balances in the world around us. In research, we are still only capable of culturing and studying only a small fraction of microbial life compared to the total diversity seen in nature, and still a far ways away from understanding all the interactions that the microbes have [6].  

**Conclusion**  

In short, humans are still a long way away from replacing microbial life, while microbial life is content to exist without humans, even if we happen to go extinct. Massive innovation and leaps in our understanding are required before consideration of replacing microbial life is even possible. This innovation would have to replace the numerous microbes, and be able capture the massive storage of information that is contained within microbes that is constantly expanding to account for the changing environment, and dynamic equilibrium state of the world's elements. In addition to this, the complex interactions that the microbes are responsible for across different environments would also need to be understood before replacement or alteration of the sensitive ecological webs can occur. If microbial life were to suddenly disappear tomorrow, or even within the next few decades, humans would almost certainly face imminent extinction. Fortunately, microbes are fairly resilient, and despite our efforts towards extinction, they are here to stay.  
    
**References** 

1. Whitman, W. B., Coleman, D. C., & Wiebe, W. J. (1998). Prokaryotes: the unseen majority. Proceedings of the National Academy of Sciences, 95(12), 6578-6583.
2. Nisbet, E. G., & Sleep, N. H. (2001). The habitat and nature of early life. Nature, 409(6823), 1083-1091.
3. Canfield D. E., Glazer A. N. & Falkowski P. G. (2010). The Evolution and Future of Earth's Nitrogen Cycle. Science, 330(192), 192-196. 
4. Chu J. (2017, September 20). Mathematics predicts a sixth mass extinction. Retrieved February 16, 2018, from http://news.mit.edu/2017/mathematics-predicts-sixth-mass-extinction-0920
5. Zak, E. (2017, February 27). 5 Invasive Species Humans Introduced on Purpose. Retrieved February 16, 2018, from https://www.care2.com/causes/5-invasive-species-humans-introduced-on-purpose.html
6. Stewart E. J. (2012). Growing Unculturable Bacteria. Journal of Bacteriology, 194(16), 4151-4160. 


## Writing assignment Module 3
*Prompt: "Discuss the challenges involved in defining a microbial species and how HGT complicates matters, especially in the context of the evolution and phylogenetic distribution of microbial metabolic pathways.  Can you comment on how HGT influences the maintenance of global biogeochemical cycles through time?  Finally, do you think it is necessary to have a clear definition of a microbial species?  Why or why not?"*

It is human nature to classify things and put things into neatly organized bins that reflect the nature and characteristics of the object, and for science and research, proper naming enables communication regarding the object in question. With organisms at a macroscopic scale, species can be defined and distinguished mostly based on whether they can interbreed. However, when it comes to microbes, there are various challenges involved in defining what a microbial species is. This is further complicated by horizontal gene transfer (HGT). In this essay, the challenges surrounding the definition of microbial species will be discussed further, along with the role of HGT in the maintenance of global biogeochemical cycles through time. Additionally, I will also discuss why it is necessary to be able to have a clear definition of what a microbial species is, even with all the challenges that come with it.  

Some of the challenges for defining a microbial species comes from the fact that many of the existing methods in use for organisms at a larger scale simply do not apply to microbes, and that existing methods are only operational. For example, the existence of asexual organisms makes it difficult to use interbreeding as a classification. Even with the use of genetic similarity to define species, including DNA hybridization and rRNA gene similarity, definitions for the cut off between what defines a species is only operational and arbitrary, and may not be reflective of actual evolutionary processes and species labelling [1]. Using this method of defining a species, there is shown to be tension between the taxonomic label, and actual similarity. For example, in the work by Thompson et al, extensive diversity was found within a group of coastal bacterioplankton, with greater than 99% 16S ribosomal RNA identity to Vibrio splendidus [2]. Specifically, they examined Hsp60 and found that it varied greatly in genome size and even carbon substrates [2]. An even greater amount of diversity was found in the work by Welch et al, in Escherichia coli, where it was shown that strains were adapted to completely different ecological niches, and only 39.2% of their combined set of proteins were common to all three strains. Whether this sort of diversity should be allowed in a single species definition is also a challenge in defining a microbial species. However, with newer approaches, such as using amplicon sequence variants (ASVs), over operational taxonomic units (OTUs) may improve this labelling [1].  

When defining a microbial species, horizontal gene transfer must also be considered, as HGT enables sharing of genes. This makes it difficult to examine evolution because evolution of microbial metabolic pathways, with HGT, may not occur linearly, as compared to other forms of events that could create genetic distance, such as mutations, insertions, and deletions. This increases the number of possible evolutionary pathways possible, and thus makes it more difficult to discern what the "true" evolutionary path is. While HGT may make phylogeny more difficult, it is an essential process in maintaining the global biogeochemical cycles. Entire metabolic pathways may be horizontally transferred, which protect the cycle, and through HGT, microbes are appropriately called the "guardians of metabolism" [3].  

Even with all of the challenges in defining a microbial species, it is essential to have one. This is because definition of a species allows for communication surrounding the group of organisms, and ensures that there is a label that is consistent between different discussions. This allows integration and findings from other studies or work to be integrated into current and future work on the group of microbes, which is most important. It would be ideal that a labelled species indicates some sort of evolutionary origin, or biological function, but for now it is most important that the labels are consistent. Consistency is even more important when it comes to a clinical setting, so that doctors may communicate to each other in a standard vocabulary, and provide proper diagnosis and treatment.  

In conclusion, while it is difficult to define a microbial species, and made even more complex with the existence of horizontal gene transfer, it is essential to have one so that discourse on the microbes can occur, and research may move forward with findings from the past. The current definition of microbial species is operational and arbitrary, and may not reflect accurately any evolutionary process or biological function, but most importantly it is consistent. Moving forward, if the definition of a microbial species does change, it would also be interesting to keep track of what organisms get moved to a different classification. As previously mentioned, consistency is most important, and if classifications do end up changing, it would also be important to consider "backwards compatibility" with the labels somehow, so that research from the past is still accessible. Hopefully, with improving methods of defining microbial species, it will be possible one day to have our definition encompass both consistency and biological meaning. 

**References**  

1. Callahan B. J., McMurdie P. J., & Holmes S. P. (2017). Exact sequence variants should replace operational taxonomic units in marker-gene data analysis. ISME, 11, 2639-2643.
2. Thompson J. R., Pacocha S., Pharino C., Klepac-Ceraj V., Hunt D. E., Benoit J., Sarma-Rupavtarm R., Distel D. L., & Polz M. F. (2005). Genotypic Diversity Within a natural Coastal Bacterioplankton population. Science, 307, 1311-1313. 
3. Falkowski P. G., Fenchel T. & Delong E. F. (2008). The Microbial Engines That Drive Earth's Biogeochemical Cycles. Science, 320, 1034-1039.



# Projects

```{r child = "MICB425_project1/MICB425_project1.Rmd"}

```

```{r child = "MICB425_project2/project2_report.Rmd"}

```